Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2021-04-01 ManuscriptTime.tex
On it

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies\replaced{ through a combination of
extrapolation, interpolation, and setting the responses for high frequencies to
zero. This translates}{, translating} into a significant reduction in
computation time. \replaced{The number of required frequencies is the most important aspect for reducing computation time. The second most important aspect is the computation grid, as t}{T}he required \deleted{computation} grids
for low frequencies (in our case in the order of 0.001\,Hz) and high
frequencies (around 100\,Hz) are hugely different. Low frequencies can be
computed on a coarser grid, but they require a much larger domain in order to
avoid boundary effects. High frequencies, on the other hand, require denser
gridding, but they are much more limited in reach. An adaptive gridding scheme
is therefore required\replaced{, which}{. We build our approach upon
\mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

@@

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{This applies primarily for expensive kernels such
as used in 3D modelling where the computation of the response of a single
frequency takes much longer than the transformation itself.} The same applies
for the reverse operation, the fewer times required, the quicker we obtain
frequency-domain results. Even though we limit our analysis here to
frequency-to-time transformations, most of the arguments are reversible and
applicable to time-to-frequency transformations. We decided to use a regular
spacing of frequencies on a log-scale, rather than an adaptive scheme. This
approach is favourable if a wide range of offsets is needed, as the required
frequencies change with offset and an adaptive frequency selection is therefore
often tailored to a single offset\added{ or a limited range of offsets}. Also,
it allows for straightforward parallelization over frequencies, which is not
completely possible with an adaptive scheme.

@@

\hrule

\added{Any Fourier transform needs a certain range of frequencies to obtain
time-domain results. Our approach is not to minimize the required for the
transform, but the number of frequencies for which we actually have to compute
responses. We do this in three zones: Above an upper threshold fmax we set the
frequency-domain response to zero. Below a lower threshold fmin we extrapolate
the frequency-domain response. As we know that the imaginary response goes
towards zero for zero frequency, we can turn it into an interpolation, for
which we use PCHIP. And in-between the thresholds we use cubic spline
interpolation in order to only have to actually compute a few distinct
frequencies. Using this approach we found that computing the responses for
15--25 frequencies is usually good enough to obtain time-domain responses for a
large range of offsets; even though the actual transform might need up to
hundreds of frequencies. The required frequencies depend naturally on the
conductivities and on the chosen acquisition type and layout. A trial-and-error
\dots}

\hrule

\replaced{The actually computed frequencies in our approach are defined by
three key-parameters}{There are three key-parameters that have to be defined}:
\replaced{The m}{M}inimum and maximum frequency ($f_\mr{min}, f_\mr{max}$) and
the number of frequencies per decade ($n_\mr{dec}^f$). \added{This creates a
frequency range which is much smaller than the frequency range required for the
transform.} \replaced{A t}{T}rial-and-error \added{approach} with a 3D code is
very time-intensive. However, a simplified, layered model for the required
survey setup and a fast 1D modeller makes it possible to estimate these
parameters easily. \replaced{An example of this is shown in
Figure~\ref{fig:FFTLog-log}.}{To this end, we created a graphical, interactive
tool using \empymod, as shown in Figure~\ref{fig:FFTLog-log}.}
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{01-FFTLog-log}
  \caption{\replaced{Example}{Interactive} frequency selection for a
    \replaced{simple}{user-provided} layered model (the shown model parameters
    are described in the text). This example shows the impulse response at an
    offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to 10\,Hz with
    five frequencies per decade. \deleted{Note that the error is clipped for
    values smaller than 0.01\,\% and bigger than 100\,\%.}}
  \label{fig:FFTLog-log}
\end{figure}
%
\replaced{It }{The shown example model} is a marine scenario with 1\,km water
depth of resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m
thick target of 100\,\ohmm at 1\,km below the seafloor in a background of
1\,\ohmm. The source is 50\,m above the seafloor, the receiver is on the
seafloor, and the response is the inline $x$-directed $E$-field. The left
subplot shows the imaginary part in the frequency domain and the right plot the
corresponding impulse response in the time domain. The red lines are the
semi-analytical responses. The blue circles indicate the actually computed
responses and the black dots the frequencies which are interpolated or set to
zero. The resulting time-domain response has a relative error of less than
1\,\% everywhere except for very early times.

Figure \ref{fig:FFTLog-lin} shows exactly the same on a
\replaced{linear}{logarithmic} scale\replaced{for the time and amplitude
axes}{, without the interactive GUI-related elements}. It can be seen that with
the chosen frequencies the time-domain response starts to divert above about
100\,s and below 0.3\,s. It also shows the oscillating high-frequency part,
which is hard to interpolate and we therefore set it to zero.
Figure~\ref{fig:DLF-lin} is the same as Figure \ref{fig:FFTLog-lin}, but
transformed with the DLF method applying the 81-point sine-cosine filter from
\cite{GEO.09.Key}. The same frequencies were computed as in the FFTLog case and
the missing ones interpolated. The error of the corresponding time-domain
response is comparable, so either FFTLog or DLF can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-FFTLog-lin}}%
  \label{fig:FFTLog-lin}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-DLF-lin}}%
  \label{fig:DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:FFTLog-log}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same frequency
  range as for the FFTLog. The resulting time-domain response has comparable
  accuracy\deleted{ (compare the error to the error in Figure
    \ref{fig:FFTLog-log})}.}
  \label{fig:lin-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. With this frequency-selection we can model a
wide range of offsets. This is shown in Figure \ref{fig:multi-offset}, where
the same kernels were used to yield the responses at offsets $r=1.5,3,6$, and
12\,km. We only need to compute 50\,\% of the required frequencies in the case
of the FFTLog, and only 14\,\% in the case of the DLF.
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:FFTLog-lin}. The coloured symbols are the actually computed
    responses, the black dots are the responses which are set to zero (high
    frequencies) or interpolated (low frequencies). The grey curves are the
    analytical responses.}
  \label{fig:multi-offset}
\end{figure*}
%

@@

\added{The proposed Fourier transformation requires a robust solver that can
compute precise results over a wide range of frequencies (times) to obtain
time-domain (frequency-domain) responses. In our case the solver uses a
stretched, regular grid and computes the electric fields in the frequency
domain. This setup is the target of the following gridding recommendations.
Whilst they will look differently for other mesh types or a time-domain code,
some conclusion will still hold.}

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the perfectly
electrically conducting boundary condition. It can also have a much higher
stretching factor $\alpha_\mr{c}$. In the presented examples we have chosen
$D_\mr{c}$ such that the distance for the signal diffusing from the source to
the boundary and back to the receiver closest to the boundary is at least two
wavelengths, after which the amplitude of the signal is reduced to a millionth
of its initial strength. The wavelength $\lambda$ to compute $D_\mr{c}$ is
given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth. Note that
for shallow marine and land cases this also applies to the horizontal
dimensions, not only to the upward $z$-direction and similarly for very
resistive basements, even in deep water. One way to circumvent this difficulty
is the use of a primary-secondary formulation, where the primary field,
including the air wave, is computed with a semi-analytical code for layered
media. We do not consider this approach here. \added{Note that grid stretching
for complex-valued diffusion fields is essentially what is done for wave fields
with the perfectly matched layer (PML). PML is an absorbing layer to avoid
scattering from the boundary by letting the field die out to zero, which is
achieved by introducing a complex factor which causes damping. As
electromagnetic diffusion fields are damped fields by themselves it suffices to
stretch the grid.}

@@

@@

\added{Note to reviewers: This entire section is new, without being all
marked in green.}

\added{mention independence of spatial complexity, why we also show 1D
examples; however, sufficiently accurate 3D results are an obvious
pre-requisite.}

@@

>>>>

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies\replaced{ through a combination of
extrapolation, interpolation, and setting the responses for high frequencies to
zero. This translates}{, translating} into a significant reduction in
computation time. \replaced{The number of required frequencies is not the only
important point for reducing computation time. Another important aspect is the computation grid, as t}{T}he required \deleted{computation} grids
for low frequencies (in our case in the order of 0.001\,Hz) and high
frequencies (around 100\,Hz) are hugely different. Low frequencies can be
computed on a coarser grid, but they require a much larger domain in order to
avoid boundary effects. High frequencies, on the other hand, require denser
gridding, but they are much more limited in reach. An adaptive gridding scheme
is therefore required\replaced{, which}{. We build our approach upon
\mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

@@

\added{Note that while we use these two algorithms for the numerical examples
any solver that provides sufficient accurate and precise frequency-domain
responses could be used.}

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{This is true in general but applies in particular
to expensive kernels such as is the case in 3D modelling, where the computation
of the response of a single frequency takes much longer than the transformation
itself.} The same applies for the reverse operation, the fewer times required,
the quicker we obtain frequency-domain results. Even though we limit our
analysis here to frequency-to-time transformations, most of the arguments are
reversible and applicable to time-to-frequency transformations. We decided to
use a regular spacing of frequencies on a log-scale, rather than an adaptive
scheme. This approach is favourable if a wide range of offsets is needed, as
the required frequencies change with offset and an adaptive frequency selection
is therefore often tailored to a single offset\added{ or a limited range of
offsets}. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

@@

\replaced{Any Fourier transform needs a certain range of frequencies to obtain
time-domain results. Our approach is not to minimize the required frequencies
for the transform, but to minimize the number of frequencies for which we
actually have to compute responses. We do this in three zones: Above an upper
threshold $f_\mr{max}$ we set the frequency-domain response to zero. Below a
lower threshold $f_\mr{min}$ we extrapolate the frequency-domain response. As
we know that the imaginary response goes towards zero for zero frequency, we
can turn it into an interpolation, for which we use PCHIP. And in-between the
thresholds we use cubic spline interpolation in order to only have to actually
compute a few distinct frequencies per decade ($n_\mr{dec}^f$). Using this
approach we found that computing the responses for 15--25 frequencies is
usually good enough to obtain time-domain responses for a large range of
offsets; even though the actual transform might need up to hundreds of
frequencies.}{There are three key-parameters that have to be defined: Minimum
and maximum frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies
per decade ($n_\mr{dec}^f$).}

\replaced{The required frequencies depend naturally on the conductivities and
on the chosen acquisition type and layout. A t}{T}rial-and-error approach with a
3D code is very time-intensive. However, a simplified, layered model for the
required survey setup and a fast 1D modeller makes it possible to estimate
these parameters easily. \replaced{An example of this is shown in
Figure~\ref{fig:FFTLog-log}.}{To this end, we created a graphical, interactive
tool using \empymod, as shown in Figure~\ref{fig:FFTLog-log}.}
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{01-FFTLog-log}
  \caption{\replaced{Example}{Interactive} frequency selection for a
    \replaced{simple}{user-provided} layered model (the shown model parameters
    are described in the text). This example shows the impulse response at an
    offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to 10\,Hz with
    five frequencies per decade. \deleted{Note that the error is clipped for
    values smaller than 0.01\,\% and bigger than 100\,\%.}}
  \label{fig:FFTLog-log}
\end{figure}
%
\replaced{It shows the responses for}{The shown example model is} a
marine scenario with 1\,km water depth of resistivity $\rho=0.3\,\ohmm$ ($\rho
= \sigma^{-1}$) and a 100\,m thick target of 100\,\ohmm at 1\,km below the
seafloor in a background of 1\,\ohmm. The source is 50\,m above the seafloor,
the receiver is on the seafloor, and the response is the inline $x$-directed
$E$-field. The left subplot shows the imaginary part in the frequency domain
and the right plot the corresponding impulse response in the time domain. The
red lines are the semi-analytical responses. The blue circles indicate the
actually computed responses and the black dots the frequencies which are
interpolated or set to zero. The resulting time-domain response has a relative
error of less than 1\,\% everywhere except for very early times.

Figure \ref{fig:FFTLog-lin} shows exactly the same on a
\replaced{linear}{logarithmic} scale \replaced{for the time and amplitude
axes}{, without the interactive GUI-related elements}. It can be seen that with
the chosen\added{, computed} frequencies \added{(blue dots)} the time-domain
response starts to divert above about 100\,s and below 0.3\,s. It also shows
the oscillating high-frequency part, which is hard to interpolate and we
therefore set it to zero. Figure~\ref{fig:DLF-lin} is the same as Figure
\ref{fig:FFTLog-lin}, but transformed with the DLF method applying the 81-point
sine-cosine filter from \cite{GEO.09.Key}. The same frequencies were computed
as in the FFTLog case and the missing ones interpolated. The error of the
corresponding time-domain response is comparable, so either FFTLog or DLF can
be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02a-FFTLog-lin}}%
  \label{fig:FFTLog-lin}}\hfill
  \subfloat[]{{\includegraphics[width=0.5\fwidth]{02b-DLF-lin}}%
  \label{fig:DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:FFTLog-log}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same frequency
  range as for the FFTLog. The resulting time-domain response has comparable
  accuracy\deleted{ (compare the error to the error in Figure
    \ref{fig:FFTLog-log})}.}
  \label{fig:lin-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. With this frequency-selection\added{, however,}
we can model a wide range of offsets. This is shown in Figure
\ref{fig:multi-offset}, where the same kernels were used to yield the responses
at offsets $r=1.5,3,6$, and 12\,km. We only need to compute 50\,\% of the
required frequencies in the case of the FFTLog, and only 14\,\% in the case of
the DLF.
%
\begin{figure*}
  \centering
  \includegraphics[width=0.75\fwidth]{03-multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:FFTLog-lin}. The coloured symbols are the actually computed
    responses, the black dots are the responses which are set to zero (high
    frequencies) or interpolated (low frequencies). The grey curves are the
    analytical responses.}
  \label{fig:multi-offset}
\end{figure*}
%

@@

\added{The proposed Fourier transformation requires a robust solver that can
compute precise results over a wide range of frequencies (times) to obtain
time-domain (frequency-domain) responses. In our case the solver uses a
regular, stretched grid and computes the electric fields in the frequency
domain. This setup is the target of the following gridding recommendations.
Whilst they will look differently for other mesh types or a time-domain code,
some conclusion will still hold.}

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the
\deleted{perfectly electrically conducting} boundary condition. It can also
have a much higher stretching factor $\alpha_\mr{c}$. In the presented examples
we have chosen $D_\mr{c}$ such that the distance for the signal diffusing from
the source to the boundary and back to the receiver closest to the boundary is
at least two wavelengths, after which the amplitude of the signal is reduced to
a millionth of its initial strength. The wavelength $\lambda$ to compute
$D_\mr{c}$ is given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth.
\deleted{Note that for shallow marine and land cases t}\added{T}his also
applies to the horizontal dimensions, not only to the upward
$z$-direction\added{,} and \replaced{it equally applies}{similarly} for very
resistive basements, even in deep water. One way to circumvent this difficulty
is the use of a primary-secondary formulation, where the primary field,
including the air wave, is computed with a semi-analytical code for layered
media. We do not consider this approach here. \added{Note that grid stretching
for complex-valued diffusion fields is essentially what is done for wave fields
with the perfectly matched layer (PML). PML is an absorbing layer to avoid
scattering from the boundary by letting the field die out to zero, which is
achieved by introducing a complex factor which causes damping. As
electromagnetic diffusion fields are damped fields by themselves it suffices to
stretch the grid.}

@@

\todo[inline]{HERE HERE HERE}

@@

\added{$\Downarrow\qquad$ Note to reviewers: The entire section 4.3 below is
new, without being all marked in green. $\qquad\Downarrow$}

\added{Mention independence of spatial complexity, why we also show 1D
examples; however, sufficiently accurate 3D results are an obvious
pre-requisite.}

@@

\added{$\Uparrow\qquad$ Note to reviewers: The entire section 4.3 above is new,
without being all marked in green. $\qquad\Uparrow$}
